### 2018-02-01 submission/manuscript.Rmd
abstract changes

**Background.** Increasing research has found that various methodological steps can have an impact on the observed microbial community when using 16S rRNA gene surveys. These components include, but are not necessarily limited to, preservation media, extraction kit, bead beating time, and primers. Both cycle number and high fidelity (HiFi) DNA polymerase are sometimes overlooked when sources of bias are being considered. Here we critically examine both cycle number and HiFi DNA polymerase for biases that may influence downstream diversity measures of 16S rRNA gene surveys.

**Methods.** DNA from Fecal samples (n = 4) were extracted using a single PowerMag DNA extraction kit with a 10 minute bead beating step and amplified at 15x, 20x, 25x, 30x, and 35x using Accuprime, Kappa, Phusion, Platinum, or Q5 HiFi DNA polymerase. Mock communities (technical replicates n = 4) consisting of previously isolated whole genomes of 8 different bacteria were also amplified using the same PCR amplification approach. First, the number of OTUs (Operational Taxonomic Units) was examined for both fecal samples and mock communites. Next, Bray-Curtis index, the error rate, sequence error prevalence, and chimera prevalence were assessed based on cycle number and HiFi DNA polymerase. Finally, the chimera prevalence correlation with number of OTUs was assessed for both cycle number and HiFi DNA polymerase dependent differences.

**Results.** At 35 cycles there were significant differences between HiFi DNA polymerase for fecal samples (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001). These HiFi dependent differences in the number of OTUs could be identified as early as 20 cycles in the mock communities (P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`). Chimera prevalence varied by HiFi DNA polymerase and these differences were still observed after chimera removal using VSEARCH. Additionally, the chimera prevalence had a strong positive correlation with the number of OTUs observed in a sample and was also not changed by chimera removal with VSEARCH.

**Conclusions.** Due to the impact of HiFi DNA polymerase on the number of OTUs, common diversity metrics that incorporate this value could give artificially inflated numbers due to higher undetected chimeras. So, when designing 16S rRNA gene survey studies it is important to consider both the cycle number and the type of HiFi DNA polymerase that will be used since it can increase or decrease the number of OTUs that are observed.

>>>>

**Background.** An increasing body of research has found that various methodological steps can have an impact on the observed microbial community when using 16S rRNA gene surveys. These components include, but are not limited to, preservation media, extraction kit, bead beating time, and primers. Both cycle number and high fidelity (HiFi) DNA polymerase are sometimes overlooked when sources of bias are being considered. Here we critically examine both cycle number and HiFi DNA polymerase for biases that may influence downstream diversity measures of 16S rRNA gene surveys.

**Methods.** DNA from Fecal samples (n = 4) were extracted using a single PowerMag DNA extraction kit with a 10 minute bead beating step and amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa, Phusion, Platinum, or Q5 HiFi DNA polymerase. Mock communities (technical replicates n = 4) consisting of previously isolated whole genomes of 8 different bacteria were also amplified using the same PCR amplification approach. First, the number of OTUs (Operational Taxonomic Units) was examined for both fecal samples and mock communites. Next, Bray-Curtis index, the error rate, sequence error prevalence, and chimera prevalence were assessed based on cycle number and HiFi DNA polymerase. Finally, the chimera prevalence correlation with number of OTUs was assessed for both cycle number and HiFi DNA polymerase dependent differences.

**Results.** At 35 cycles there were significant differences between HiFi DNA polymerase for fecal samples (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001). These HiFi dependent differences in the number of OTUs could be identified as early as 20 cycles in the mock communities (P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`). Chimera prevalence varied by HiFi DNA polymerase and these differences were still observed after chimera removal using VSEARCH. Additionally, the chimera prevalence had a strong positive correlation with the number of OTUs observed in a sample and this association was not changed by chimera removal with VSEARCH.

**Conclusions.** Due to the impact of HiFi DNA polymerase on the number of OTUs, common diversity metrics that incorporate this value could give artificially inflated numbers due to higher undetected chimeras. When designing 16S rRNA gene survey studies it is important to consider both the cycle number and the type of HiFi DNA polymerase that will be used since it can increase or decrease the number of OTUs that are observed.

### 2018-02-01 submission/manuscript.tex
abstract changes

\textbf{Background.} Increasing research has found that various
methodological steps can have an impact on the observed microbial
community when using 16S rRNA gene surveys. These components include,
but are not necessarily limited to, preservation media, extraction kit,
bead beating time, and primers. Both cycle number and high fidelity
(HiFi) DNA polymerase are sometimes overlooked when sources of bias are
being considered. Here we critically examine both cycle number and HiFi
DNA polymerase for biases that may influence downstream diversity
measures of 16S rRNA gene surveys.

\textbf{Methods.} DNA from Fecal samples (n = 4) were extracted using a
single PowerMag DNA extraction kit with a 10 minute bead beating step
and amplified at 15x, 20x, 25x, 30x, and 35x using Accuprime, Kappa,
Phusion, Platinum, or Q5 HiFi DNA polymerase. Mock communities
(technical replicates n = 4) consisting of previously isolated whole
genomes of 8 different bacteria were also amplified using the same PCR
amplification approach. First, the number of OTUs (Operational Taxonomic
Units) was examined for both fecal samples and mock communites. Next,
Bray-Curtis index, the error rate, sequence error prevalence, and
chimera prevalence were assessed based on cycle number and HiFi DNA
polymerase. Finally, the chimera prevalence correlation with number of
OTUs was assessed for both cycle number and HiFi DNA polymerase
dependent differences.

\textbf{Results.} At 35 cycles there were significant differences
between HiFi DNA polymerase for fecal samples (P-value \textless{}
0.0001). These HiFi dependent differences in the number of OTUs could be
identified as early as 20 cycles in the mock communities (P-value =
0.002). Chimera prevalence varied by HiFi DNA polymerase and these
differences were still observed after chimera removal using VSEARCH.
Additionally, the chimera prevalence had a strong positive correlation
with the number of OTUs observed in a sample and was also not changed by
chimera removal with VSEARCH.

\textbf{Conclusions.} Due to the impact of HiFi DNA polymerase on the
number of OTUs, common diversity metrics that incorporate this value
could give artificially inflated numbers due to higher undetected
chimeras. So, when designing 16S rRNA gene survey studies it is
important to consider both the cycle number and the type of HiFi DNA
polymerase that will be used since it can increase or decrease the
number of OTUs that are observed.

>>>>

\textbf{Background.} An increasing body of research has found that
various methodological steps can have an impact on the observed
microbial community when using 16S rRNA gene surveys. These components
include, but are not limited to, preservation media, extraction kit,
bead beating time, and primers. Both cycle number and high fidelity
(HiFi) DNA polymerase are sometimes overlooked when sources of bias are
being considered. Here we critically examine both cycle number and HiFi
DNA polymerase for biases that may influence downstream diversity
measures of 16S rRNA gene surveys.

\textbf{Methods.} DNA from Fecal samples (n = 4) were extracted using a
single PowerMag DNA extraction kit with a 10 minute bead beating step
and amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa,
Phusion, Platinum, or Q5 HiFi DNA polymerase. Mock communities
(technical replicates n = 4) consisting of previously isolated whole
genomes of 8 different bacteria were also amplified using the same PCR
amplification approach. First, the number of OTUs (Operational Taxonomic
Units) was examined for both fecal samples and mock communites. Next,
Bray-Curtis index, the error rate, sequence error prevalence, and
chimera prevalence were assessed based on cycle number and HiFi DNA
polymerase. Finally, the chimera prevalence correlation with number of
OTUs was assessed for both cycle number and HiFi DNA polymerase
dependent differences.

\textbf{Results.} At 35 cycles there were significant differences
between HiFi DNA polymerase for fecal samples (P-value \textless{}
0.0001). These HiFi dependent differences in the number of OTUs could be
identified as early as 20 cycles in the mock communities (P-value =
0.002). Chimera prevalence varied by HiFi DNA polymerase and these
differences were still observed after chimera removal using VSEARCH.
Additionally, the chimera prevalence had a strong positive correlation
with the number of OTUs observed in a sample and this association was
not changed by chimera removal with VSEARCH.

\textbf{Conclusions.} Due to the impact of HiFi DNA polymerase on the
number of OTUs, common diversity metrics that incorporate this value
could give artificially inflated numbers due to higher undetected
chimeras. When designing 16S rRNA gene survey studies it is important to
consider both the cycle number and the type of HiFi DNA polymerase that
will be used since it can increase or decrease the number of OTUs that
are observed.
